---
title: "Final Project"
author:
  - Lihui Chen
  - Xuechen Yu
  - Richard Li (Xuanyou)
output: rmarkdown::html_vignette
# output: pdf_document
vignette: >
  %\VignetteIndexEntry{Finalproj}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Background and Motivation
Metastatic colorectal cancer (mCRC), which is a significant challenge in oncology, remains something that has a substantial impact on patient morbidity and mortality. Even though there have been a lot of advancements in therapeutic challenges, the prognosis for patients with mCRC is often poor. The selection of first-line therapy is thus a critical determinant of clinical outcomes. 

The management of mCRC has evolved with the introduction of targeted therapies, which aim to improve survival rates and quality of life. Among these, the epidermal growth factor receptor (EGFR) inhibitor panitumumab has shown promise when used in combination with the conventional chemotherapy regimen FOLFOX (a combination of folinic acid, fluorouracil, and oxaliplatin).

The trial, NCT00364013, is a randomized, multicenter, phase 3 study. It was designed to evaluate the efficacy of panitumumab in combination with FOLFOX versus FOLFOX alone as first-line therapy for patients with previously untreated mCRC. This study holds a repository of rich clinical data, including patient demographics, treatment details, response criteria, survival metrics, and adverse events. Such comprehensive data presents an invaluable opportunity for the application of advanced statistical and machine learning models to predict patient outcomes.

The motivation behind utilizing these computational models lies in their ability to investigate complex patterns within the data. It potentially leads to the identification of prognostic variables and the development of predictive models for patient survival. By applying these methodologies, the research can shed light on the multifaceted nature of mCRC progression and response to treatment. Moreover, it strives to enhance the decision-making process in clinical settings, enabling personalized medicine approaches and optimizing therapeutic efficacy.

# 2. Research question
1. How can robust statistical and machine learning models be developed and validated using the trial dataset to accurately predict mortality in patients with metastatic colorectal cancer (mCRC)?

2. What are the key clinical and treatment-related factors that significantly impact the survival of patients with mCRC, and how can their influence be quantified and incorporated into predictive models for patient survival?

# 3. Data cleaning and exploration

The data are separated into a number of files, and we primarily focused on the adsl and biomark data. We first converted the data type and then imported the file into R studio. The adsl file is full_joined by biomark file, and then we converted the categorical data using one hot encoding and dummy variable. Except the categorial column of sex and race, all the other categorical variables are converted using one hot encoding. 

In the data preprocessing phase of our analysis, we focused on a selected set of variables believed to influence patient outcomes in metastatic colorectal cancer. We first isolated these key variables and then quantified the missing values for each. To maintain the integrity of our analysis, we opted for listwise deletion, removing any records with missing data to prevent potential biases that could arise from incomplete information. This resulted in a refined dataset, which was cleansed of missing values and thus poised for the next stage of our investigation. We documented the reduction in the dataset size to ensure transparency in our methodology and the implications it may have on the study's findings.

```{r setup}
library(bis620.2022)
library(readr)
library(dplyr)
library(tidyr)
```

```{r}
data <- adsl %>%
  full_join(biomark, by = 'SUBJID')

final_df_clean = cleaning_helperfunc(data)
help(cleaning_helperfunc)
# Description
# This function performs several cleaning operations on a data frame. It converts categorical columns to numeric, focusing on specific variables of interest, and removes rows with missing values. It also provides a summary of missing values in the specified variables of interest.
```

```{r}
head(final_df_clean)
```

# 4. Analysis
```{r}
# Relevant Package Installation
suppressPackageStartupMessages(library(randomForest))
suppressPackageStartupMessages(library(rpart.plot))
suppressPackageStartupMessages(library(caTools))
suppressPackageStartupMessages(library(caret))
suppressPackageStartupMessages(library(pROC))
```

## 4.1 Survival Analysis -- Coxph Model
```{r}
suppressPackageStartupMessages(library(survival))
Model1 = fit_coxph(final_df_clean)
?fit_coxph
# Description
# This function fits a Cox Proportional Hazards model to the provided dataset using specified covariates. It is designed to analyze survival data, specifically to model the time until an event of interest (or hazard) occurs.
```

```{r}
Model1
```

## 4.2 Machine Learning Model1 -- Logistic Regression

```{r}
suppressPackageStartupMessages(library(caTools))
suppressPackageStartupMessages(library(caret))
suppressPackageStartupMessages(library(pROC))

# Splitting the data into training and test sets
train_test_split <- function(df){
  
  set.seed(123)
  split <- sample.split(final_df_clean$DTH, SplitRatio = 0.7)
  train_data <- final_df_clean[split == TRUE, ]
  test_data <- final_df_clean[split == FALSE, ]
  
  return(list(train = train_data, test = test_data))
}

data_all = train_test_split(final_df_clean)
train_data = data_all$train
test_data = data_all$test

Model2 = fit_logistic_regression(train_data, test_data)
?fit_logistic_regression
# Description:
# Fits a logistic regression model to the training data and evaluates its performance on the test data. The function prints the model's accuracy and Area Under the Curve (AUC) for the Receiver Operating Characteristic (ROC) curve. It also returns the fitted model object.
```

```{r}
Model2
```

## 4.3 Machine Learning Model2 -- Random Forest

```{r}
suppressPackageStartupMessages(library(randomForest))
suppressPackageStartupMessages(library(rpart.plot))

# we defined a new function called fit random forest
# The following is our function description:
# Fits a random forest model to the provided training data and evaluates its performance on the test data. It prints the model's accuracy and generates a ROC curve with the AUC value. The function also plots variable importance and returns the fitted random forest model object.
?fit_random_forest
```

```{r}
Model3 = fit_random_forest(train_data, test_data)
```

## 4.4 Visualization

```{r}
generate_roc_for_two <- function(logistic_roc, rf_roc, logistic_auc, rf_auc){
    
  par(pty = "s")
  plot(logistic_roc, col="#377eb8", lwd = 2.5, main="ROC Curves Comparison")
  # Add ROC curve for random forest
  lines(rf_roc, col="#4daf4a", lwd = 2.5)
  
  legend("bottomright", legend=c("Logistic Regression", "Random Forest"), 
         col=c("#377eb8", "#4daf4a"), lty=c(1, 2), cex=0.7)
  
  # Optionally, add the AUC scores to the legend if you have them calculated
  # Assume logistic_auc_value and rf_auc_value hold the AUC values for logistic regression and random forest, respectively
  legend("bottomright", legend=c(paste("Logistic Regression AUC =", round(logistic_auc, 4)), 
                                 paste("Random Forest AUC =", round(rf_auc, 4))), 
         col=c("#377eb8", "#4daf4a"), lty=c(1, 2), cex=0.7)
  
  par(pty = "m")

}

generate_roc_for_two(roc_curve, rf_roc_curve, logistic_auc_value, rf_auc_value)
```
# 5. Interpretation and conclusions
